Direct-style process creation on Linux

### 2020-05-04 research/paper/paper.tex
DEV REVISE partial stuff

\subsection{Example: Create and exec}
\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process and immediately execing}]
# Use the clone syscall in the local thread to create a new thread;
# we use a wrapper that supplies defaults for all arguments.
child = local.thread.clone()
# Call execve to run a different executable in the child thread;
# We pass the executable path as the first argument in the argument list, as is traditional.
# We use a wrapper that defaults envp to an unchanged environment, so we don't pass envp.
child.execve(hello_path, [hello_path])
\end{lstlisting}
In Listing \ref{basic}, we simply create a new process under our control (a thread)
and immediately exec a binary with it.
As exceptions are used for error-handling in the Python API,
there is no need for error-checking code.
%% **** note to self: concepts introduced in this example
%% local.thread
%% clone
%% execve
\subsection{Example: Passing down FDs}
\begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
child = local.thread.clone()
sock = child.socket(AF.INET, SOCK.DGRAM)
# Bind the socket to a sockaddr_in;
# the sockaddr is allocated in memory with child.ptr and is garbage collected.
sock.bind(child.ptr(SockaddrIn(0, 0)))
sock.listen(10)
sock.disable_cloexec()
child.execve(executable_path, [executable_path, "--listening-socket", str(int(sock))])
\end{lstlisting}
In Listing \ref{fds}, we create a new thread,
then create a listening socket bound to a random port in that thread,
then call exec, 
passing down the socket by disabling cloexec and passing its file descriptor number as an argument to the new program.

File descriptors, here, are object oriented and have relevant syscalls as methods.
They make syscalls in the process they are created in by default;
we can create more objects referring to the same file descriptor from different processes
if we want to make the syscalls from another process.
%% **** note to self: concepts introduced in this example
%% disable_cloexec
%% socket creation
%% using int on sock
%% .ptr
\subsection{Example: Pipeline}
\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a pipeline}]
# create the pipe
pipe = local.thread.pipe()
child1 = local.thread.clone()
# inherit the write-end of the pipe to child1, and replace child1.stdout with it
child1.inherit_fd(pipe.write).dup2(child1.stdout)
child1_proc = child1.execve(yes_path, [yes_path])
child2 = local.thread.clone()
# inherit the read-end of the pipe to child2, and replace child2.stdin with it
child2.inherit_fd(pipe.read).dup2(child2.stdin)
child2_proc = child2.execve(head_path, [head_path, "-n", "15"])
\end{lstlisting}
In Listing \ref{pipe},
we do the same as the Unix shell pipeline "yes | head -n 15".
We create a pipe,
then create two threads,
connect them with a pipe,
and exec a different program in each thread.

@@

Then we simply =dup2= as normal to replace child1's stdout with the write end of the pipe;
=dup2= disables CLOEXEC by default on the target.
%% **** note to self: concepts introduced in this example
%% dup2
%% pipe
\subsection{Example: Mount namespace}
\begin{lstlisting}[float,language=Python,label={mount},caption={Mount namespace}]
child = local.thread.clone(CLONE.NEWUSER|CLONE.NEWNS)
child.mkdir(rootdir/"proc")
child.mount(Path("/proc"), rootdir/"proc", "", MS.BIND, "")
child.chroot(rootdir)
child.execve(executable_path, [executable_path])
\end{lstlisting}
In Listing \ref{mount},
we make a new mount namespace and rearrange the filesystem tree for the child process.
We bind-mount /proc at /proc inside the chroot directory,
chroot into the directory,
and exec an executable which will run inside the chroot.
%% **** note to self: concepts introduced in this example
%% namespaces
%% path slash syntax sugar
\subsection{Example: Nested clone and network namespace}
\begin{lstlisting}[float,language=Python,label={nested},caption={Nested clone and network namespace}]
ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)

listening_child = ns_thread.clone()
sock = listening_child.socket(AF.INET, SOCK.DGRAM)
sockaddr = SockaddrIn(22, "127.0.0.1")
sock.bind(listening_child.ptr(sockaddr))
sock.listen(10)
sock.disable_cloexec()
child.execve(server_path, [server_path, "--listening-socket", str(int(sock))])

connecting_child = ns_thread.clone()
child.execve(client_path, [
  client_path, "--connect-address", str(sockaddr.address) + ":" + str(sockaddr.port)])
\end{lstlisting}
In Listing \ref{nested},
we make a process (\texttt{ns\_thread}) in a new network namespace.
Then, we create two more child processes of \texttt{ns\_thread},
which are also in the new network namespace.
This nested creation of child processes is fully supported,
like all other syscalls,
and allows us to set up complex graphs of processes and namespaces.

We bind to a privileged port on localhost inside the namespace,
and create one child to listen on that socket,
and another child to connect to it.
%% **** note to self: concepts introduced in this example
%% network namespace
%% nested clone
% \subsection{Example: Starting a complex system}
% \begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
% ### create socket outside network namespace that Miredo will use for internet access
% inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
% inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
% # set some miscellaneous additional sockopts that Miredo wants
% set_miredo_sockopts(local.thread, inet_sock)
% ### create main network namespace thread
% ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
% ### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
% icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
% ### create in-network-namespace socket,
% ### which Miredo will use for unassociated Ifreq ioctls
% reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
% ### create and set up the TUN interface
% tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
% ### create socketpair which Miredo will use to communicate
% ### between privileged process and Teredo client
% privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
% ### start up privileged process which manipulates the network setup in the namespace
% privproc_thread = ns_thread.clone()
% # We preserve NET_ADMIN capability over exec so that privproc can
% # manipulate the TUN interface. add_to_ambient_caps is a helper function
% # used because manipulating Linux ambient capabilities is fairly verbose
% add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
% # privproc expects to communicate with the main client over stdin and stdout
% privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
% privproc_side.dup2(privproc_thread.stdin)
% privproc_side.dup2(privproc_thread.stdout)
% privproc_child = privproc_thread.execve(miredo_privproc_executable_path, [
%     miredo_privproc_executable_path, str(tun_index)
% ])
% ### start up Miredo client process which communicates
% ### over the internet to implement the tunnel
% # the client process doesn't need to be in the same network namespace,
% # since it is passed all the resources it needs as fds at startup.
% client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
% # lightly sandbox by unmounting everything except for the executable
% # and its deps (known via package manager)
% unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
% # a helper function for preparing the fds that are passed as command line arguments
% async def pass_fd(fd: FileDescriptor) -> str:
%     client_thread.inherit_fd(fd).disable_cloexec()
%     return str(int(fd))
% client_child = client_thread.execve(miredo_client_executable_path, [
%     miredo_client_executable_path,
%     pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
%     pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
%     "teredo.remlab.net", "teredo.remlab.net"
% ])
% \end{lstlisting}

@@

Rewrite some existing system with rsyscall?
Write a shell?
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.

>>>>

\subsection{Example: Create and exec}
\begin{lstlisting}[float,language=Python,label={basic},caption={Creating a new process and immediately execing}]
# Use the clone syscall in the local thread to create a new thread;
# we use a wrapper that supplies defaults for all arguments.
child = local.thread.clone()
# Call execv to run a different executable in the child thread;
# We pass the executable path as the first argument in the argument list, as is traditional.
child.execv(hello_path, [hello_path])
\end{lstlisting}
In Listing \ref{basic}, we simply create a new process under our control (a thread)
and immediately exec a binary with it.
As exceptions are used for error-handling in the Python API,
there is no need for error-checking code.
%% **** note to self: concepts introduced in this example
%% local.thread
%% clone
%% execve
\subsection{Example: Passing down FDs}
\begin{lstlisting}[float,language=Python,label={fds},caption={Passing down FDs}]
child = local.thread.clone()
sock = child.socket(AF.INET, SOCK.DGRAM)
# Bind the socket to a sockaddr_in;
# the sockaddr is allocated in memory with child.ptr and is garbage collected.
sock.bind(child.ptr(SockaddrIn(0, 0)))
sock.listen(10)
sock.disable_cloexec()
child.execv(executable_path, [executable_path, "--listening-socket", str(int(sock))])
\end{lstlisting}
In Listing \ref{fds}, we create a new thread,
then create a listening socket bound to a random port in that thread,
then call exec, 
passing down the socket by disabling cloexec and passing its file descriptor number as an argument to the new program.

File descriptors, here, are object oriented and have relevant syscalls as methods.
They make syscalls in the process they are created in by default;
we can create more objects referring to the same file descriptor from different processes
if we want to make the syscalls from another process.
%% **** note to self: concepts introduced in this example
%% disable_cloexec
%% socket creation
%% using int on sock
%% .ptr
\subsection{Example: Pipeline}
\begin{lstlisting}[float,language=Python,label={pipe},caption={Creating a pipeline}]
# create the pipe
pipe = local.thread.pipe()
child1 = local.thread.clone()
# inherit the write-end of the pipe to child1, and replace child1.stdout with it
child1.inherit_fd(pipe.write).dup2(child1.stdout)
child1_proc = child1.execv(yes_path, [yes_path])
child2 = local.thread.clone()
# inherit the read-end of the pipe to child2, and replace child2.stdin with it
child2.inherit_fd(pipe.read).dup2(child2.stdin)
child2_proc = child2.execv(head_path, [head_path, "-n", "15"])
\end{lstlisting}
In Listing \ref{pipe},
we do the same as the Unix shell pipeline "yes | head -n 15".
We create a pipe,
then create two threads,
connect them with a pipe,
and exec a different program in each thread.

@@

Then we simply =dup2= as normal to replace child1's stdout with the write end of the pipe;
=dup2= disables CLOEXEC by default on the target.
%% **** note to self: concepts introduced in this example
%% dup2
%% pipe
\subsection{Example: Mount namespace}
\begin{lstlisting}[float,language=Python,label={mount},caption={Mount namespace}]
child = local.thread.clone(CLONE.NEWUSER|CLONE.NEWNS)
child.mkdir(rootdir/"proc")
child.mount(Path("/proc"), rootdir/"proc", "", MS.BIND, "")
child.chroot(rootdir)
child.execv(executable_path, [executable_path])
\end{lstlisting}
In Listing \ref{mount},
we make a new mount namespace and rearrange the filesystem tree for the child process.
We bind-mount /proc at /proc inside the chroot directory,
chroot into the directory,
and exec an executable which will run inside the chroot.
%% **** note to self: concepts introduced in this example
%% namespaces
%% path slash syntax sugar
\subsection{Example: Nested clone and network namespace}
\begin{lstlisting}[float,language=Python,label={nested},caption={Nested clone and network namespace}]
ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)

listening_child = ns_thread.clone()
sock = listening_child.socket(AF.INET, SOCK.DGRAM)
sockaddr = SockaddrIn(22, "127.0.0.1")
sock.bind(listening_child.ptr(sockaddr))
sock.listen(10)
sock.disable_cloexec()
child.execv(server_path, [server_path, "--listening-socket", str(int(sock))])

connecting_child = ns_thread.clone()
child.execv(client_path, [
  client_path, "--connect-address", str(sockaddr.address) + ":" + str(sockaddr.port)])
\end{lstlisting}
In Listing \ref{nested},
we make a process (\texttt{ns\_thread}) in a new network namespace.
Then, we create two more child processes of \texttt{ns\_thread},
which are also in the new network namespace.
This nested creation of child processes is fully supported,
like all other syscalls,
and allows us to set up complex graphs of processes and namespaces.

We bind to a privileged port on localhost inside the namespace,
and create one child to listen on that socket,
and another child to connect to it.
%% **** note to self: concepts introduced in this example
%% network namespace
%% nested clone
% \subsection{Example: Starting a complex system}
% \begin{lstlisting}[float,language=Python,label={miredo},caption={Starting a complex system}]
% ### create socket outside network namespace that Miredo will use for internet access
% inet_sock = local.thread.socket(AF.INET, SOCK.DGRAM)
% inet_sock.bind(local.thread.ptr(SockaddrIn(0, 0)))
% # set some miscellaneous additional sockopts that Miredo wants
% set_miredo_sockopts(local.thread, inet_sock)
% ### create main network namespace thread
% ns_thread = local.thread.clone(CLONE.NEWNET|CLONE.NEWUSER)
% ### create in-network-namespace raw INET6 socket which Miredo will use to relay pings
% icmp6_fd = ns_thread.socket(AF.INET6, SOCK.RAW, IPPROTO.ICMPV6)
% ### create in-network-namespace socket,
% ### which Miredo will use for unassociated Ifreq ioctls
% reqsock = ns_thread.socket(AF.INET, SOCK.STREAM)
% ### create and set up the TUN interface
% tun_fd, tun_index = make_tun(ns_thread, "miredo", reqsock)
% ### create socketpair which Miredo will use to communicate
% ### between privileged process and Teredo client
% privproc_pair = ns_thread.socketpair(AF.UNIX, SOCK.STREAM)
% ### start up privileged process which manipulates the network setup in the namespace
% privproc_thread = ns_thread.clone()
% # We preserve NET_ADMIN capability over exec so that privproc can
% # manipulate the TUN interface. add_to_ambient_caps is a helper function
% # used because manipulating Linux ambient capabilities is fairly verbose
% add_to_ambient_caps(privproc_thread, {CAP.NET_ADMIN})
% # privproc expects to communicate with the main client over stdin and stdout
% privproc_side = privproc_thread.inherit_fd(privproc_pair.first)
% privproc_side.dup2(privproc_thread.stdin)
% privproc_side.dup2(privproc_thread.stdout)
% privproc_child = privproc_thread.execv(miredo_privproc_executable_path, [
%     miredo_privproc_executable_path, str(tun_index)
% ])
% ### start up Miredo client process which communicates
% ### over the internet to implement the tunnel
% # the client process doesn't need to be in the same network namespace,
% # since it is passed all the resources it needs as fds at startup.
% client_thread = ns_thread.clone(CLONE.NEWUSER|CLONE.NEWNET|CLONE.NEWNS|CLONE.NEWPID)
% # lightly sandbox by unmounting everything except for the executable
% # and its deps (known via package manager)
% unmount_everything_except(client_thread, miredo_exec.run_client.executable_path)
% # a helper function for preparing the fds that are passed as command line arguments
% async def pass_fd(fd: FileDescriptor) -> str:
%     client_thread.inherit_fd(fd).disable_cloexec()
%     return str(int(fd))
% client_child = client_thread.execv(miredo_client_executable_path, [
%     miredo_client_executable_path,
%     pass_fd(inet_sock), pass_fd(tun_fd), pass_fd(reqsock),
%     pass_fd(icmp6_fd), pass_fd(privproc_pair.second),
%     "teredo.remlab.net", "teredo.remlab.net"
% ])
% \end{lstlisting}

@@

Rewrite some existing system with rsyscall?
Write a shell?

Right okay so we can do some fork-style code, sure.
Show the IPC, not too complex, but show it.

We can do it in Python with rsyscall - have a hidden experimental secret fork thing.
Using write_all_bytes will be annoying but fine.
Ugh, but we'll have to decode it too.
Eh, no, we can wrap that.
Yeah then it's fine.
\section{Future work}\label{future_work}
\subsection{Other applications of rsyscall}
Most avenues of future work focus on rsyscall.
rsyscall was not developed solely for the purpose of this paper,
and it has many uses unrelated to direct-style process creation,
such as asynchronous system calls, exceptionless system calls\cite{flexsc}, cross-host operations, among others.
We are actively exploring such applications,
as well as broadening rsyscall's language support.
\subsection{Kernel support}
rsyscall's cross-process syscalls can be performed entirely in userspace,
which has substantial benefits for deployability.
Nevertheless, direct kernel support for creating a stub process and performing syscalls in the context of that process
may provide efficiency benefits, as well as reducing userspace-visible complexity.
